---
title: "MDSR Ch. 3 Exercises" 
author: "TJ Schaeffer"
date: "Due: January 20, 2019"
output: html_notebook
---


```{r Front Matter, include=FALSE}
### Note: `include=FALSE` causes R to run the code, but does not show this chunk in the R Notebook
# always clean up R environment
rm(list = ls())

# packages
library(ggplot2)
library(mdsr)
library(tidyverse)
library(lubridate)
library(mosaic)
library(mosaicData)
library(NHANES)
library(macleish)
library(babynames)
library(Hmisc)
library(graphics)

# inputs & source data
data("NHANES")
data("babynames")
```

## Figure 3.18
```{r}
# Inspect the Data
NHANES

# Data Wrangling
NHANES2 <-
  NHANES %>%
  select(AgeDecade, BMI_WHO, Diabetes) %>%
  arrange(AgeDecade)

NHANES2Clean <-
  NHANES2[- c(1:2766), ]

NHANES2Plot <-
  mosaicplot(~ AgeDecade + BMI_WHO + Diabetes, data = NHANES2, shade = TRUE, xlab = "AgeDecade")
```

## Figure 3.27

# part A
# Inspect the data
babynames

# Data wrangling
BabynamesDist <- make_babynames_dist()

# com_fem <- 
  BabynamesDist %>% 
  filter(sex == "F") %>% 
  group_by(name) %>% 
  summarise( 
    N = n(), 
    est_num_alive = sum(est_alive_today), 
    q1_age = wtd.quantile(age_today, est_alive_today, probs = 0.25),
    median_age = wtd.quantile(age_today, est_alive_today, probs = 0.5),
    q3_age = wtd.quantile(age_today, est_alive_today, probs = 0.75)) %>%
arrange(desc(est_num_alive)) %>%
  head(25)

w_plot <-
  ggplot(data = com_fem,
         aes(x = reorder(name, -median_age), y = median_age)) +
  xlab(NULL) +
  ylab("Age (in years)") +
  ggtitle("Median ages for females with the 25 most common names") 

w_plot <-
  w_plot +
  geom_linerange(aes(ymin = q1_age, ymax = q3_age), color = "#f3d478", size = 10, alpha = 0.8)

w_plot <-
  w_plot +
  geom_point(fill = "#ed3324", colour = "white", size = 4, shape = 21)

w_plot +
  geom_point(aes(y = 55, x = 24), fill = "#ed3324", colour = "white", size = 4, shape = 21) +
  geom_text(aes(y = 58, x = 24, label = "median")) +
  geom_text(aes(y = 26, x = 16, label = "25th")) +
  geom_text(aes(y = 51, x = 16, label = "75th percentile")) +
  geom_point(aes(y = 24, x = 16), shape = 17) +
  geom_point(aes(y = 56, x = 16), shape = 17) +
  coord_flip()

```{r}
# part B
# Inspect the data
babynames

# Data wrangling
BabynamesDist <- make_babynames_dist()

com_fem <- 
  BabynamesDist %>% 
  filter(sex == "F", est_alive_today > 0) %>% #Add "est_alive_today > 0" to get rid of NA error
  group_by(name) %>% 
  summarise( 
    N = n(), 
    est_num_alive = sum(est_alive_today), 
    q1_age = wtd.quantile(age_today, est_alive_today, probs = 0.25),
    median_age = wtd.quantile(age_today, est_alive_today, probs = 0.5),
    q3_age = wtd.quantile(age_today, est_alive_today, probs = 0.75)) %>%
arrange(desc(est_num_alive)) %>%
  head(25)

w_plot <-
  ggplot(data = com_fem,
         aes(x = reorder(name, -median_age), y = median_age)) +
  xlab(NULL) +
  ylab("Age (in years)") +
  labs(title = "Median ages for females with the 25 most common names", # Changed "ggtitle"" function to "labs" in order to add subtitle and caption
       subtitle = "Among Americans to be alive as of Jan. 1, 2014",
       caption = "Source: Social Security Administration")

w_plot <-
  w_plot +
  geom_linerange(aes(ymin = q1_age, ymax = q3_age), color = "#f3d478", size = 10, alpha = 0.8)

w_plot <-
  w_plot +
  geom_point(fill = "#ed3324", colour = "white", size = 4, shape = 21)

w_plot +
  geom_point(aes(y = 55, x = 24), fill = "#ed3324", colour = "white", size = 4, shape = 21) +
  geom_text(aes(y = 58, x = 24, label = "median")) +
  geom_text(aes(y = 26, x = 16, label = "25th")) +
  geom_text(aes(y = 51, x = 16, label = "75th percentile")) +
  geom_point(aes(y = 24, x = 16), shape = 17) +
  geom_point(aes(y = 56, x = 16), shape = 17) +
  ylim(15, 75) + # Added y limit
  coord_flip()
```
* Added "est_alive_today > 0" to get rid of NA error
* Changed "ggtitle"" function to "labs" in order to add subtitle and caption
* Added y limit
























