---
title: "MDSR Chapter 07 Exercises"
author: "TJ Schaeffer"
date: "Due: 0/03/2019"
output: html_notebook
---

# Front matter

```{r}
# always clean up R environment
rm(list = ls())

# load all packages here
library(ggplot2)
library(mdsr)
library(tidyverse)
library(lubridate)
library(mosaic)
library(mosaicData)
library(NHANES)
library(macleish)
library(babynames)
library(Hmisc)
library(Lahman)
library(readr)
library(xtable)
library(devtools)
library(rvest)
# user-defined functions here (if any)

# load data
data("Gestation") # from mosaicData
```


# Exercises

### Exercise 7.1
```{r}
# p. 168
Bootstrap71 <- 
  mosaic::do(500) * mean(~ age, data = sample_n(Gestation, size =200, replace = TRUE), na.rm = TRUE)
confint(Bootstrap71, level = .95)
```

** We are 95% confident that the mean age of mothers lies between 27 and 28 years old.**

### Exercise 7.2
```{r}
Bootstrap72 <- 
  mosaic::do(500) * median(~ age, data = sample_n(Gestation, size =200, replace = TRUE), na.rm = TRUE)
confint(Bootstrap72, level = .95)
```

** We are 95% confident that the median age of mothers lies between 26 and 27 years old.**

### Exercise 7.3
```{r}
Bootstrap73 <- 
  mosaic::do(500) * lm(wt ~ age, data = sample_n(Gestation, size =200, replace = TRUE), na.rm = TRUE)
confint(Bootstrap73, level = .95)
```



### Exercise 7.4
** Since the empirical rule says that standard deviations multiplier at 95% confidence interval is 2 and the standard deviations multiplier at a 99.7% confidence interval is 3, that means that the standarad deviations multiplier at the 99% confidence interval is between 2 and 3. After looking it up on the interent, the amount of standard deviations that should be used at a 99% confidence interval is 2.58.



### Excercise 7.5
The model says that as wind conditions increase/worsen, the distance in the flight of the ball increases. However, the R-squared value is only 9%, which means that the model is very weak.


### Excercise 7.6
The second model includes more variables and is better than the first model because the R-squared value is strong at 59%. Additionally, the Cond plays has a smaller p-value and therefore has a stronger impact on the Dist variable than in the first model. However, the second model is not profoundly better because some of the variables have a high p-value, which means that they don't have a very strong impact on the Dist variable. Overall, in order to make the second model better, you must take out the variables with high p-values. 
